---
title: "Case Study 2 DDS"
author: "Nicholas Sager"
date: "2023-04-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Introduction text

For an interactive app to visualize this data, please see: [Shiny App](https://nicksager.shinyapps.io/CaseStudy1/)

## Data Description

A description of the data, where it came from, and what it contains.

### Read the Data

First, we will read in the beers data, and merge the breweries data so that each beer has brewery data associated with it. Then we will view a few row to make sure it looks ok. 


## Data Exploration

### NA Values

### Visualizing Variables of interest

Salary and attrition
- summary statistics
- relationship to eachother


### Salary and Attrition by other metrics (department / job title)

boxplots

### Pairwise scatterplots of key variables


## Modelling

### Transformations and further cleaning

### Attrition

sensitivity + specificity (>60 + >60)

#### KNN

#### Naive Bayes

### Salary

RMSE (<$3000)

#### Linear Regression
Include assumptions

#### Logistic regression?
Maybe include if there is more time

## Results and Discussion

Top 3 Factors predicting attrition

Other Tests

## Conclusion

Conclusion text

This analysis has answered the initial questions posed by Budweiser's request, and has perhaps raised even more. The author hopes that this analysis will be useful to Budweiser in their future product development. Any questions about this analysis or proposals for additional research can be directed to the author at:

Nicholas Sager: nsager@smu.edu  

## Appendix 

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```



